Efficient simulation of individual-based population models

### 2024-03-26 article.qmd
insurance portfolio

>>>>

This section provides an example of how to use the `IBMPopSim` package
to simulate a heterogeneous life insurance portfolio (see
also `vignette(IBMPopSim_insurance_portfolio)`).

We consider an insurance portfolio consisting of male policyholders, of
age greater than 65. These policyholders are characterized by their age,
assumed to be less than $a_{max} = 110$, and risk class
$x\in \mathcal X =\{1,2\}$.

**Entries in the portfolio** New policyholders enter the population at a
constant Poisson rate $\mu^{en}=\lambda$, which means that on average,
$\lambda$ individuals enter the portfolio each year. A new individual
enters the population at an age a that is uniformly distributed between
65 and 70, and is in risk class 1 with probability $p$.

**Death events** A baseline age and time specific death rate is first
calibrated on "England and Wales (EW)" males mortality historic
data[^1], and projected for 30 years using the Lee-Carter model with the
package `StMoMo` (see [@stmomo]). The forecasted baseline death
intensity is denoted by $d(t,a)$, defined by:
$$
d(t,a) = \sum_{k=0}^{29}\mathbf{1}_{\{k\leq t < k+1\}} d_k(a), \quad \forall \; t\in [0,30] \text{ and } a \in [65, a_{max}],
$$ {#eq-insurance-baseline}
with $d_k(a)$ the point estimate of the forecasted mortality rate for
age $a$ and year $k$.\
Individuals in risk class 1 are assumed to have mortality rates that are
20% higher than the baseline mortality (for instance, the risk class
could refer to smokers), while individuals in risk class 2 are assumed
to have mortality rates that are 20% lower than the baseline (non
smokers). The death intensity of an individual $I= (\tau_b, \infty, x)$,
of age $a(I,t) = t - \tau_b$ at time $t$ and in risk class
$x \in \{1, 2\}$ is thus the function $$\label{eq-insurance-deathrates}
\lambda^d(t,I) = \alpha_x d(t,a(I,t)), \quad \alpha_1 = 1.2, \quad \alpha_2 = 0.8.$$
In particular, the death intensity verifies Assumption TODO
[2](#AssumptionIntensity1){reference-type="ref"
reference="AssumptionIntensity1"} since:
$$
\lambda^d(t,I) \leq \bar d : = \alpha_1 \sup_{t \in [0,30]} d(t,a_{max}).
$$ {#eq-insurance-bound-deathrates}

**Exits from the portfolio** Individuals exit the portfolio at a
constant (individual) rate $\lambda^{ex}(t,I) = \mu^{i}$ only depending
on their risk class $i\in \{1,2\}$.

## Population {#insurance-population}

We start with an initial population of $30\,000$ males of age 65,
distributed uniformly in each risk class. The population data frame has
thus the two (mandatory) columns `birth` (here the initial time is
$t_0=0$) and `death` (`NA` if alive), and an additional column
`risk_cls` corresponding to the policyholders risk class. Since there
are entry and exit events, the `entry` and `out` flags of the population
constructor are set to `TRUE`.

```{r}
N <- 30000
pop_df <- data.frame("birth" = rep(-65,N), "death" = rep(NA,N),
                     "risk_cls" = rep(1:2,each=N/2))
pop_init <- population(pop_df, entry=TRUE, out=TRUE)
```

## Events {#insurance-events}

#### Entry events

The age of the new individual is determined by the `kernel_code`
argument in the `mk_event_poisson` function.

```{r}
entry_params <- list("lambda" = 30000, "p" = 0.5)
entry_event <- mk_event_poisson(
    type = "entry",
    intensity = "lambda",
    kernel_code = "if (CUnif() < p) newI.risk_cls =1;
                   else newI.risk_cls= 2;
                   double a = CUnif(65, 70);
                   newI.set_age(a, t);")
```

Note that the variables `newI` and `t`, as well as the function
`CUnif()`, are implicitly defined and usable in the `kernel_code`. The
field `risk_cls` comes from the names of characteristics of individuals
in the population. The names `lambda` and `p` are parameter names that
will be specified in the `R` named list `params`.

Here we use a constant $\lambda$ as the event intensity, but we could
also use a rate $\lambda(t)$ that depends on time, using the function
`mk_event_poisson_inhomogeneous`.

#### Death and exit events

The baseline death intensity defined
in @eq-insurance-baseline and obtained with the package
`StMoMo` is stored in the variable `death_male`.

```{r}
# StMoMo death rates
library('StMoMo')
library('reshape2')
EWStMoMoMale <- StMoMoData(EWdata_hmd, series = "male")
LC <- lc()
ages.fit <- 65:100
years.fit <- 1950:2016
LCfitMale <- fit(LC, data = EWStMoMoMale, ages.fit = ages.fit, years.fit = years.fit)
t <- 30
LCforecastMale <- forecast(LCfitMale, h = t)
d_k <- apply(LCforecastMale$rates, 2, function(x) stepfun(66:100, x))
breaks <- 1:29
death_male <- piecewise_xy(breaks,d_k)

death_max <- max(sapply(d_k, function(x) { max(x) }))

death_params <- list("death_male" = death_male, "alpha" = c(1.2, 0.8))
death_event <- mk_event_individual(
    type = "death",
    intensity_code = "result = alpha[I.risk_cls-1] * death_male(t, I.age(t));")
```

The death and exit intensities are of class `individual` (see Table
@tbl-intensity-classes). Hence, the death and exit events
are created with the `mk_event_individual` function.

```{r}
death_params <- list("death_male" = death_male, "alpha" = c(1.2, 0.8))
death_event <- mk_event_individual(
    type = "death",
    intensity_code = "result = alpha[I.risk_cls-1] * death_male(t, I.age(t));")
```

```{r}
exit_params = list("mu" = c(0.001, 0.06))
exit_event <- mk_event_individual(
    type = "exit",
    intensity_code = "result = mu[I.risk_cls-1]; ")
```

## Model creation and simulation {#insurance-simulation}

The model is created from all the previously defined building blocks, by
calling the `mk_model`.

```{r}
model <- mk_model(
    characteristics = get_characteristics(pop_init),
    events = list(entry_event, death_event, exit_event),
    parameters = c(entry_params, death_params, exit_params))
```

Once the model is compiled, it can be used with different parameters and
run simulations for various scenarios. Similarly, the initial population
(here `pop_df`) can be modified without rerunning the `mk_model`
function. The bounds for entry events is simply the intensity $\lambda$.
For death events, the bound is given by $\bar{d}$ defined in
@eq-insurance-bound-deathrates, which is stored in the
`death_max` variable.

```{r}
bounds <- c("entry" = entry_params$lambda,
            "death" = death_max,
            "exit" = max(exit_params$mu))

sim_out <- popsim(
    model = model,
    initial_population = pop_init,
    events_bounds = bounds,
    parameters = c(entry_params, death_params, exit_params),
    time = 30,
    age_max = 110,
    multithreading = TRUE)
```

## Outputs

The data frame `sim_out$population` consists of all individuals present
in the portfolio during the period of $[0, 30]$, including the
individuals in the initial population and those who entered the
portfolio. Each row represents an individual, with their date of birth,
date of death (`NA` if still alive at the end of the simulation), risk
class, and characteristics `entry` and `out`. Recall that if an
individual enters the population at time $t$, his `entry` characteristic
is automatically set up to be equal to $t$. The characteristics `out` is
set to `TRUE` for individuals who left the portfolio due to an exit
event.

In this example, the simulation time over 30 years, starting from an
initial population of 30 000 individuals is of $2\times 10^{-4}$
seconds, for an acceptance rate of proposed event of approximately 25%.
At the end of the simulation, the number of alive individuals is
approximately 430 000.

Initially in the portfolio (at $t=0$), there is the same number of 65
years old policyholders in each risk class. However, policyholders in
the risk class 2 with lower mortality rates leave the portfolio at
higher rate than policyholders in the risk class 1 : $\mu^2 > \mu^1$.
Therefore, the heterogeneous portfolio composition changes with time,
including more and more individuals in risk class 1 with higher
mortality rates, but with variations across age classes. To illustrate
the composition of the total population at the end of the simulation
($t=30$), we present in Figure @fig-insur-a the age pyramid of the final composition of
the portfolio obtained with the `age_pyramid` and `plot` function of the
`pyramid` class.

`IBMPopSim` also allows the fast computation of exact life tables from
truncated and censored individual data (due to entry and exit events),
using the functions `death_table` and `exposure_table`. These function
are particularly efficient, since the computations are made using the
`Rccp` library.

In Figure @fig-insur-b, we illustrate the central death rates in the
simulated portfolio at final time. Due to the mortality differential
between risk class 1 and 2, one would expect to observe more individuals
in risk class 2 at higher ages. However, due to exit events, a higher
proportion of individuals in risk class 1 exit the portfolio over time,
resulting in a greater proportion of individuals in risk class 1 at
higher ages than what would be expected in the absence of exit events.
Consequently, the mortality rates in the portfolio are more aligned with
those of risk class 1 at higher ages. This is a simple example of how
composition changes in the portfolio can impact aggregated mortality
rates and potentially compensate or reduce an overall mortality
reduction (see also [@KAAKAI201916]).

```{r, echo=FALSE, fig.width=10}
#| label: fig-insur-a
#| fig-cap: "Portfolio age pyramid at t = 30 for individuals in risk class 1 (blue) and 2 (red)."
#| fig-pos: 'h'

dim(population_alive(sim_out$population,t = 30))

sim_out$logs[["duration_ns"]]/1e9

sim_out$logs[["effective_events"]] / sim_out$logs[["proposed_events"]]

# Age pyramid (Figure 2(a))

age_grp <- 65:98
pyr = age_pyramid(sim_out$population, time = 30, ages=age_grp)
colnames(pyr)[2]<- "group_name"
pyr$group_name <- as.character(pyr$group_name)
colors <- c("1"="#00AFBB","2"="#FC4E07")
plot(pyr,colors,age_breaks = as.integer(seq(1,length(age_grp)-1,by=2)))
```

```{r, echo=FALSE, fig.width=10}
#| label: fig-insur-b
#| fig-cap: "Portfolio central death rates at t = 30 (black)."
#| fig-pos: 'h'

# Portofolio death rates

age_grp <- 65:95
Dx_pop <- death_table(sim_out$population, ages = age_grp, period = 0:30)
Ex_pop <- exposure_table(sim_out$population, ages = age_grp, period = 0:30)
mx_pop <- Dx_pop/Ex_pop

# Figure 2(b)

Dx <- death_table(sim_out$population[sim_out$population$risk_cls==1,],
                  ages = age_grp, period = 0:30)
Ex <- exposure_table(sim_out$population[sim_out$population$risk_cls==1,],
                     ages = age_grp, period = 0:30)

LC <- lc()
LCfitSim1 <- fit(LC, Dxt = Dx , Ext = Ex,ages=age_grp[-length(age_grp)])
time <- 30
age <- age_grp[-length(age_grp)]
estimated <- LCfitSim1$ax + LCfitSim1$bx*LCfitSim1$kt[time]
forecast_1 <- log(1.3*LCforecastMale$rates[1:(length(age_grp)-1),time])
forecast_2 <- log(0.8*LCforecastMale$rates[1:(length(age_grp)-1),time])
portfolio <- log((mx_pop)[,time])

df <- data.frame(age=age, portfolio=portfolio, forecast_1=forecast_1, forecast_1=forecast_2)
colnames(df) <- c("Age", "Portfolio", "Risk class 1", "Risk class 2")
df_melt <- melt(df, id="Age")

ggplot(data=df_melt, aes(x=Age, y=value)) +
  geom_point(aes(color=variable, shape=variable)) +
  geom_line(aes(color=variable, linetype=variable)) +
  xlab("Age") +
  ylab("Log mortality rates") +
  scale_color_manual(values = c("Portfolio" = "black", "Risk class 1" = "blue", "Risk class 2" = "red")) +
  scale_shape_manual(values = c("Portfolio" = 1, "Risk class 1" = NA, "Risk class 2" = NA)) +
  scale_linetype_manual(values = c("Portfolio" = 0, "Risk class 1" = 2, "Risk class 2" = 1)) +
  theme(legend.title = element_blank(), legend.position = c(0.9,0.2), plot.title = element_text(hjust = 0.5))
